---
title: "Predict Future Sales"
author: "Pep Porrà"
date: "March, 5th 2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Read Data

(https://www.kaggle.com/c/competitive-data-science-predict-future-sales/overview)

```{r}
list.files("data/")
```

Read in Russian item, category and shop descriptions

```{r}
#Sys.setlocale()
Sys.setlocale(locale = "Russian")
Sys.setlocale(locale = "English")
#Sys.setlocale(locale = "")
```


```{r}
load("data/dataset_all.RData")
load("data/testset_all.RData")
load("data/items.RData")
load("data/item_categories.RData")
load("data/shops.RData")
```


## Feature checks and properties


Check shop and item categories numbers. Test OK

```{r}
length(unique(dataset_all$item_id)) == max(items$item_id) + 1
length(unique(dataset_all$item_id)) <= max(items$item_id) + 1
length(unique(dataset_all$item_category_id)) == max(item_categories$item_category_id) + 1
length(unique(dataset_all$shop_id)) == max(shops$shop_id) + 1

```

Less items in the train_sales than in items. Probably there are items in the test set that do not exist in the train set (363). It will be challenging to forecast these products.

```{r}
length(unique(testset_all$item_id))
length(unique(testset_all$shop_id))
length(unique(testset_all$item_id)) * length(unique(testset_all$shop_id))
sum(!unique(testset_all$item_id) %in% unique(dataset_all$item_id) )
sum(!items$item_id %in% unique(dataset_all$item_id))
```

In the testset, all shops (42 in total) can sell the same total number of item types (5100). So the size of the testset is `r 42 * 5100`.


check NA

```{r}
sum(is.na(dataset_all))
```

```{r}
summary(dataset_all)
```

Check if prices changed. Conclusion: prices can change probably by shop and/or date

```{r}
vec_sd <- tapply(dataset_all$item_price, dataset_all$item_id, sd)
sum(is.na(vec_sd))
names(vec_sd) <- NULL
sum(vec_sd[!is.na(vec_sd)] <= 0.00000000000000000000001)
all.equal(tapply(dataset_all$item_price, dataset_all$item_id, sd), rep(0, 21807))
```

Trend per day

```{r}
q_by_day <- tapply(dataset_all$item_cnt_day, dataset_all$date, sum)
```


```{r}
df_q_by_day <- data.frame(date = as.Date(names(q_by_day)), q = as.numeric(q_by_day))
plot(df_q_by_day$date, df_q_by_day$q, log = "y", typ = "l",
  las = 1, cex.axis = 0.7,
  ylab = "Quantity", xlab ="Day")
abline(v= seq.Date(from = as.Date("2013-01-01"), 
  to = as.Date("2015-10-31"), by = "3 month"), lty = c(2), col = "gray")
grid(nx = NA, ny = NULL)
```

weekly seasonality

```{r}
plot(df_q_by_day$q ~ factor(weekdays(df_q_by_day$date, abbreviate = T),
  levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
  xlab = "Weekday", ylab = "quantity")
grid()
```



Q by Month

```{r}
q_by_month <- tapply(dataset_all$item_cnt_day, dataset_all$date_block_num, sum)
```


```{r}
df_q_by_month <- data.frame(month = 0:33, q = as.numeric(q_by_month))
plot(df_q_by_month$month, df_q_by_month$q, log = "y",
  las = 1, cex.axis = 0.7,
  ylab = "Quantity", xlab = "Month index")
abline(v= c(0, 6, 12, 18, 24, 30), lty = c(2, 3), col = "gray")
grid(nx = NA, ny = NULL)
```

Q by shop

```{r}
q_by_shop <- tapply(dataset_all$item_cnt_day, dataset_all$shop_id, sum)
```


```{r}
df_q_by_shop <- data.frame(shop = 0:59, q = as.numeric(q_by_shop))
plot(df_q_by_shop$shop, df_q_by_shop$q, log = "y", las = 1,
  ylab = "Quantity", xlab = "shop id", cex.axis = 0.7)
grid()

```

Number of different items sold by shop

```{r}
unique_by_shop <- tapply(dataset_all$item_id, dataset_all$shop_id, function(x) length(unique(x)))
```


```{r}
df_unique_by_shop <- data.frame(shop = 0:59, q = as.numeric(unique_by_shop))
plot(df_unique_by_shop$shop, df_unique_by_shop$q, las = 1,
  ylab = "number of distinct items", xlab = "Shop id", cex.axis = 0.7)
grid()
abline(h = 5100, col = "red", lty = 2)
```

```{r}
sum(dataset_all$item_cnt_day == 0)
```
We can assume all shops can sell the same items and that when the quantity is zero it is not registered.

Returns

```{r}
sum(dataset_all$item_cnt_day < 0)
sum(dataset_all$item_cnt_day > 0)
```

Sales by category

```{r}
q_by_category <- tapply(dataset_all$item_cnt_day, dataset_all$item_category_id, sum)
```


```{r}
df_q_by_category <- data.frame(category = 0:83, q = as.numeric(q_by_category))
plot(df_q_by_category$category, df_q_by_category$q, log = "y",
  ylab = "Quantity of items", xlab = "Category id", las = 1, cex.axis = 0.8)
grid()
```


Number of unique shops per month


```{r}
unique_shops_by_month <- tapply(dataset_all$shop_id, dataset_all$date_block_num, 
  function(x) length(unique(x)))
```


```{r}
df_unique_shops_by_month <- data.frame(month = 0:33, q = as.numeric(unique_shops_by_month))
plot(df_unique_shops_by_month$month, df_unique_shops_by_month$q,
  ylab = "Quantity of shops", xlab = "Month id", las = 1, cex.axis = 0.8,
  xlim = c(0, 34))
# Add shops in the test dataset
test_df <- data.frame(month = 34, q = length(unique(testset_all$shop_id)))
points(test_df$month, test_df$q, pch = 16, col = "red")
grid()
```

Is there any new shop in the test month?

```{r}
sum(!unique(testset_all$shop_id) %in% unique(dataset_all[dataset_all$date_block_num == 33, ]$shop_id))
```

All shops in the test month existed also in the month before. This is a test to run before using any model

Compute number of shops-items per month

```{r}
unique_shop_item_by_month <- tapply(dataset_all$shop_item, dataset_all$date_block_num, 
  function(x) length(unique(x)))
```


```{r}
df_unique_shop_item_by_month <- data.frame(month = 0:33, q = as.numeric(unique_shop_item_by_month))
plot(df_unique_shop_item_by_month$month, df_unique_shop_item_by_month$q,
  ylab = "Quantity of shops_item", xlab = "Month id", las = 1, cex.axis = 0.8,
  xlim = c(0, 34), ylim = c(0, 250000))
# Add shops in the test dataset
test_df <- data.frame(month = 34, q = nrow(testset_all))
points(test_df$month, test_df$q, pch = 16, col = "red")
grid()
```

Number of months active per shop

```{r}
q_months_by_shop <- tapply(dataset_all$date_block_num, dataset_all$shop_id, 
  function(x) length(unique(x)))
```


```{r}
df_q_months_by_shop <- data.frame(shop = 0:59, q = as.numeric(q_months_by_shop))
plot(df_q_months_by_shop$shop, df_q_months_by_shop$q,
  ylab = "Quantity of months active", xlab = "Shop id", las = 1, cex.axis = 0.8,
  xlim = c(0, 59))
# Add shops in the test dataset
test_df <- data.frame(shop = 0:59, q = (0:59) %in% unique(testset_all$shop_id))
points(df_q_months_by_shop[test_df$q, ]$shop, 
  df_q_months_by_shop[test_df$q, ]$q, pch = 16, col = "red")
grid()
legend("bottomright", legend = c("shop in the test set"), col = "red", pch = 16, cex = 0.7)
```


Shop with less than 34 months that are included in the testset

```{r}
shops_in_set <- test_df[test_df$q, ]$shop
df_q_months_by_shop[
  (df_q_months_by_shop$shop %in% shops_in_set) & (df_q_months_by_shop$q <34), ]$shop
```

## Model for a given month

Let us forecast month, where n between 0 and 33.
We forget for a moment about daily and wekkly and monthly seasonalities. Our model is that in month n we will sell the same as in month n-1.

```{r}
load("data/df_month.RData")
```


```{r}
sum(df_month$item_cnt_day == 0)
sum(df_month$item_cnt_day > 0)
sum(df_month$item_cnt_day < 0)
max(df_month$item_cnt_day)
```


```{r}
q_different <- table(df_month$item_cnt_day[df_month$item_cnt_day > 0])
```


```{r}
head(q_different)
tail(q_different)
```

```{r}
df_month[df_month$item_cnt_day == max(df_month$item_cnt_day), ]
```


```{r}
range(q_different)
range(df_month$item_cnt_day)
```


```{r}
plot(as.numeric(names(q_different)), q_different, log = "xy", ylim = c(1, 2e6),
  yaxt = "n",
  xlab = "Quantity of an item sold", ylab = "#items")
axis(2, at = 10^(0:6), labels = c(1, 10, 100, 1000, 10000, 100000, 1000000), las = 1)

fit_qq <- lm(log(q_different, 10) ~ log(as.numeric(names(q_different)), 10) )
abline(fit_qq, col = "red", lty = 2)
abline(h = c(1, 10, 100, 1000, 10000, 100000, 1000000),
  v = c(1, 5, 10, 50, 100, 500, 1000), lty = 3, col = "lightgray")
```

## Explore intermediate months

```{r}
month <- 23
month_set <- dataset_all[dataset_all$date_block_num == month, ]
all_combinations <- expand.grid(shops = unique(month_set$shop_id), item = unique(month_set$item_id))
```

```{r}
cat("Distinct shop_item in month", month, ":", nrow(month_set), "\n")
cat("potential shop_item in month", month, ":", nrow(all_combinations), "\n")
```


```{r}
all_shop_item_id <- paste(all_combinations$shops, all_combinations$item, sep = "_")
```

```{r}
all_combinations$shop_item <- all_shop_item_id
all_combinations$q <- 0
# merge(dataset_all[dataset_all$date_block_num == month - 1, c("shop_item", "item_count_day")])
```

We could consider the dataset formed by all possibles items (22170), in all shops (60) and for all months (34) or days

```{r}
22170 * 60
```

This is the number of time series we want to forecast for the month Nov-2015. In October 2015, the total number of shop-items was

```{r}
length(unique(df_month[ df_month$date_block_num == 33, ]$shop_item))
```

these are the values dictinct of zero. The total numbers shops-items would be

```{r}
length(unique(df_month[ df_month$date_block_num == 33, ]$shop_id))
length(unique(df_month[ df_month$date_block_num == 33, ]$item_id))
length(unique(df_month[ df_month$date_block_num == 33, ]$shop_id)) * 
  length(unique(df_month[ df_month$date_block_num == 33, ]$item_id))

```

In Nov-2015, we have `r nrow(testset_all)`. Total number of shop_item in month 34 that does not exist in month 33 are

```{r}
sum(!unique(testset_all$shop_item) %in% unique(df_month[ df_month$date_block_num == 33, ]$shop_item))
```

But if we check shops first

```{r}
cat("Distinct shops in month 33:", length(unique(testset_all$shop_id)), "\n")
cat("Distinct shops in month 34:", 
  length(unique(df_month[ df_month$date_block_num == 33, ]$shop_id)), "\n")
cat("Difference:", 
  sum(!unique(testset_all$shop_id) %in% unique(df_month[ df_month$date_block_num == 33, ]$shop_id)),
  "\n")

```

and items later

```{r}
cat("Distinct items in month 33:", length(unique(testset_all$item_id)), "\n")
cat("Distinct items in month 34:", 
  length(unique(df_month[ df_month$date_block_num == 33, ]$item_id)), "\n")
cat("Difference:", 
  sum(!unique(testset_all$item_id) %in% unique(df_month[ df_month$date_block_num == 33, ]$item_id) ),
  "\n")
```

```{r}
5100 * 42
```

There are 1109 items which (we have to assume) have a quantity sold greater than 0 that were not sold the previous month. Let us check if this situation happens in previous months

```{r}
items_not_in_previous_month <- function(n){
  stopifnot(n > 0 & n <= 34)
  ifelse(n == 34,
    items_n <- unique(testset_all$item_id),
    items_n <- unique(df_month[ df_month$date_block_num == n, ]$item_id))
  items_previous_month <- 
    unique(df_month[ df_month$date_block_num == n - 1, ]$item_id)
  difference <- !items_n %in% items_previous_month
  q <- sum(df_month[ 
    df_month$date_block_num == n &
    df_month$item_id %in% items_n[difference] , ]$item_cnt_day)
  return(c(sum(difference), q))
}
```

```{r}
m_not_in_previous_month <- t(sapply(1:34, items_not_in_previous_month))
colnames(m_not_in_previous_month) <- c("#items", "q_sold")
m_not_in_previous_month
```


```{r}
items_new <- function(n){
  stopifnot(n > 0 & n <= 34)
  ifelse(n == 34,
    items_n <- unique(testset_all$item_id),
    items_n <- unique(df_month[ df_month$date_block_num == n, ]$item_id))
  items_previous_month <- 
    unique(df_month[ df_month$date_block_num <= n - 1, ]$item_id)
  difference <- !items_n %in% items_previous_month
  q <- sum(df_month[ 
    df_month$date_block_num == n &
    df_month$item_id %in% items_n[difference] , ]$item_cnt_day)
  return(c(sum(difference), q))
}
```

Every month there are new items introduced. The forecast of these items is the challenging part of this problem
Potentially we can separate the items in 3 categories:
* sold month before
* sold before but not the month before
* new items


```{r}
m_new <- t(sapply(1:34, items_new))
colnames(m_new) <- c("#items", "q_sold")
m_new
```

```{r}
items_month <- function(n){
  stopifnot(n >= 0 & n <= 34)
  ifelse(n == 34,
    items_n <- unique(testset_all$item_id),
    items_n <- unique(df_month[ df_month$date_block_num == n, ]$item_id))
  q <- sum(df_month[df_month$date_block_num == n, ]$item_cnt_day)
  return(c(length(items_n), q))
}
```


```{r}
items_month(34)
```

```{r}
m_items <- t(sapply(0:34, items_month))
colnames(m_items) <- c("#items", "q_sold")
m_items
```

```{r}
dates <- seq.Date(as.Date("2013-01-01"), as.Date("2015-11-01"), by = "month")
plot(dates, m_items[, 1], xlab = "Months",
  ylab = "Number of different references (items)", las = 1, xaxt = "n")
axis.Date(1, dates)
x_ticks = axis.Date(1, dates, labels = F)

grid(nx = NA, ny = NULL)
abline(v = x_ticks, col ="lightgray", lty = 3)
points(as.Date("2015-11-01"), m_items[35, 1], col = "red", pch = 16)
```



```{r}
m_not_in_previous_month - m_new
```

```{r}
items_in_previous_month <- function(n){
  stopifnot(n > 0 & n <= 34)
  ifelse(n == 34,
    items_n <- unique(testset_all$item_id),
    items_n <- unique(df_month[ df_month$date_block_num == n, ]$item_id))
  items_previous_month <- 
    unique(df_month[ df_month$date_block_num == n - 1, ]$item_id)
  difference <- items_n %in% items_previous_month
  q <- sum(df_month[ 
    df_month$date_block_num == n &
    df_month$item_id %in% items_n[difference] , ]$item_cnt_day)
  return(c(sum(difference), q))
}
```

```{r}
m_in_previous_month <- t(sapply(1:34, items_in_previous_month))
colnames(m_in_previous_month) <- c("#items", "q_sold")
m_in_previous_month
```

```{r}
m_not_in_previous_month + m_in_previous_month
```

```{r}
sum(m_items[-1, ] != m_in_previous_month + (m_not_in_previous_month - m_new) + m_new)
```

```{r}
month <- 23
cat("items in month (", month, "):", m_items[month + 1, ], "\n")
cat("items in previous month (", month - 1, "):", m_in_previous_month[month, ], "\n")
cat("items not in previous month (", month - 1, ") but in some month before:", 
  m_not_in_previous_month[month, ] - m_new[month, ], "\n")
cat("items new in month (", month , "):", m_new[month, ], "\n")
```




